Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-01-27 analysis/paper/paper.Rmd
edits on discussion

Our CV tests indicates that there are significant differences in shape standardisation between the pre-European period and post-European period, and between the Chinese contact and either pre-European or post-European periods. This suggests pottery shape became more homogeneous and standardized after foreign contacts with European colonizers and even moreso after contact with Chinese immigrants. Compositional analysis shows that the clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources. We can thus rule out changes in clay fabric as a factor in changes in vessel shape. To determine if shape changes might be related to changes in the function of pots at Kiwulan, we attempted to identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay <!-- add citation of work that we are trying to follow here, eg. Seung's paper in JIPA -->. Unfortunately, we did not obtain useful results due to very low lipid yields, which might result from the very thin, dense, and low porosity fabric of rhte Kiwulan pottery. These physical characteristics of the clay offers limited spaces to trap and protect organic molecules from microbiological degradation [cf. @Evershed2008, pp. 909].

Our results support the hypothesis that foreign influences at Kiwulan influenced emergence of social inequality in the local Indigenous society. We find that , stronger in the Chinese period instead of the European period originally thought to be. The hypothesis of the emergence of social inequality based on the monopolies for long-distance trade brought by the Europeans was not fully supported. The nature and context of the interaction with the foreign presences might give an insight into the different degrees of standardized pottery. Compared to other region in Taiwan, the European colonial control was weak in Yilan due to the relatively isolated location by mountains and the economic consideration of the Spanish and the Dutch who prefer northern Taiwan as a trading base. Yilan could be seen as a place experienced indirect influences by the European trade network and their colonial activities in a pericolonial context [cf. @Acabado2017]. On the contrary, the interaction between indigenous people and the Chinese in the 19th century was more intense and direct since those Chinese immigrants settled in Yilan and lived closely with indigenous societies. This direct influence is reflected by the archaeological evidence of a large amount of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009]. Also, the finding of adoption of coffin in mortuary practice that viewed as an ethnic symbol for Chinese was observed [@Chen2007].

>>>>

Our CV tests indicates that there are significant differences in shape standardisation between the pre-European period and post-European period, and between the Chinese contact and either pre-European or post-European periods. This suggests pottery shape became more homogeneous and standardized after foreign contacts with European colonizers and even more so after contact with Chinese immigrants. Compositional analysis shows that the clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources. We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape. To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay <!-- add citation of work that we are trying to follow here, eg. Seung's paper in JIPA -->. Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery. These physical characteristics of the clay offers limited spaces to trap and protect organic molecules from microbiological degradation [cf. @Evershed2008, pp. 909].

Our results support the hypothesis that foreign influences at Kiwulan influenced emergence of social inequality in the local Indigenous society. We find that foreign influence may have been stronger in the Chinese period than the European period, contrary to previous work <!-- add citation to previous work -->. The hypothesis of an emergence of social inequality based on the monopolies for long-distance trade brought by the Europeans was not fully supported. <!-- add more detail about why this hypothesis was not fully supported --> Compared to other regions in Taiwan, European colonial influence was weak in Yilan due to the isolated of the surrounding mountains and the economic focus of the Spanish and  Dutch who prefered northern Taiwan as their trading base. Indigenous communities Yilan experienced indirect influence from European trade networks and their colonial activities in a pericolonial context [cf. @Acabado2017]. In contrast to the Indigenous-European interactions at Kiwulan, the interaction between indigenous people and  Chinese in the 19th century appears to have been more intense and direct. Historical records indicate that  Chinese immigrants settled in Yilan and lived closely with Kiwulan Indigenous societies <!-- add citation that is the source of this information --> . This direct influence is reflected by the archaeological evidence of large amounts of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009]. Similarly, burials at Kiwulan in this later phase show the adoption of coffins in mortuary practices, which @Chen2007 interprets as the adoption of a symbol of ethnic Chinese [@Chen2007].
